```{r}
library(readxl)
library(dplyr)
library(ggplot2)
library(vegan)
library(glmmTMB)
library(DHARMa)
library(effects)
library(emmeans)
library(openxlsx)
library(car)
```


```{r}
#Import the bird data set using the read.csv command####
#birds_2014 <- read.csv("C:/Users/Lenovo/Documents/BTR/Bird_Data_New/test.csv")
birds_2015 <- read.csv("C:/Users/Lenovo/Documents/BTR/Bird_Data_New/STOC-EPS-2015-ASSEMBLE.csv")
birds_2016 <- read.csv("C:/Users/Lenovo/Documents/BTR/Bird_Data_New/STOC-EPS-2016_assemble.csv")
birds_2017 <- read.csv("C:/Users/Lenovo/Documents/BTR/Bird_Data_New/STOC-EPS-2017_assemble.csv")
birds_2018 <- read.csv("C:/Users/Lenovo/Documents/BTR/Bird_Data_New/STOC-EPS-2018_assemble.csv")
birds_2019 <- read.csv("C:/Users/Lenovo/Documents/BTR/Bird_Data_New/STOC-EPS-2019-ASSEMBLE.csv")
birds_2021 <- read.csv("C:/Users/Lenovo/Documents/BTR/Bird_Data_New/STOC-EPS-2021.csv")
birds_2024 <- read.csv("C:/Users/Lenovo/Documents/BTR/Bird_Data_New/STOC-EPS-2024-ASSEMBLE.csv")

#Remove all extra columns void of information####
#Compares left of %in% to the right, if matching, because of "!", the code returns it as FALSE and removes it from the dataframe.
birds_2015 <- birds_2015[, !(names(birds_2015) %in% paste0("X.", 15))]
birds_2016 <- birds_2016[, !(names(birds_2016) %in% paste0("X.", 15))]
birds_2017 <- birds_2017[, !(names(birds_2017) %in% paste0("X.", 15))]
birds_2018 <- birds_2018[, !(names(birds_2018) %in% paste0("X.", 15))]
birds_2019 <- birds_2019[, !(names(birds_2019) %in% paste0("X.", 15))]
birds_2021 <- birds_2021[, !(names(birds_2021) %in% paste0("X.", 15))]
birds_2024 <- birds_2024[, !(names(birds_2024) %in% paste0("X.", 15))]

#REMOVING TRANSIT DATA####
#Create a function, create a loop, apply that function and save each year in a list
remove_obs <- function(df, column) {
  df[!df[[column]] %in% c("TOTAL", "Vol", "en chemin", ">200m","Vol (<200m)", ""),] #"Vol (<200m)" and "" or smth
} #df is the dataframe, column is the columns the name of the column to filter by. Matches on both sides of the %in% and removes it due to the "!".

#Create a list of all your yearly datasets, "data" is the name of the dataset, "col" is the desired column you would like to filter
birds_list <- list(
  #birds_2014 = list(data = birds_2014, col = "X.14"),
  birds_2015 = list(data = birds_2015, col = "X.14"),
  birds_2016 = list(data = birds_2016, col = "X.14"),
  birds_2017 = list(data = birds_2017, col = "X.14"),
  birds_2018 = list(data = birds_2018, col = "X.14"),
  birds_2019 = list(data = birds_2019, col = "X.14"),
  birds_2021 = list(data = birds_2021, col = "X.14"),
  birds_2024 = list(data = birds_2024, col = "X.14")
)

for(i in names(birds_list)){
  birds_list[[i]]$data <- remove_obs(birds_list[[i]]$data, birds_list[[i]]$col)
}

#names(birds_list) → gives the names of each list element: "birds_2014", "birds_2015", etc.
  #for(i in names(birds_list)) → loops over each dataset.
    #birds_list[[i]]$data → selects the dataframe for the current year.
      #birds_list[[i]]$col → selects the column name to filter by.
        #remove_obs(...) → applies the filtering function.
          #<- → replaces the old dataframe with the cleaned version.

new_colnames <- c(
  "Subsite","Site","Type (Restauration active / passive / control)","julian_day",
  "Date","Heure debut","1er ou 2e passage",
  "cloudiness", "rain", "wind", "visibility", "surface_water",
  "Observer Name","Department","Distance"
)

birds_list <- lapply(birds_list, function(x) {
  names(x$data)[1:length(new_colnames)] <- new_colnames
  return(x)
})

head(birds_list$birds_2015$data)
```

```{r}
#Removing repeated header rows#### 

#1. Extract the cleaned dataframes from the list
cleaned_dfs <- lapply(birds_list, function(x) x$data)

cleaned_dfs <- lapply(cleaned_dfs, function(df) {
  df[-1, ]  # remove the first row
})

#2. Identify all columns across all years
all_cols <- unique(unlist(lapply(cleaned_dfs, names)))

#3. Ensure all dataframes have the same columns
# If a column doesn’t exist in a year, fill with NA
standardize_cols <- function(df, all_cols) {
  missing_cols <- setdiff(all_cols, names(df))
  if(length(missing_cols) > 0){
    df[missing_cols] <- NA  # add missing columns
  }
  df <- df[all_cols]        # reorder columns consistently
  return(df)
}
cleaned_dfs <- lapply(cleaned_dfs, standardize_cols, all_cols = all_cols)

#4. Combine all years into a single dataset
birds_all_years <- do.call(rbind, cleaned_dfs)

birds_all_years$Subsite <- gsub("Platelet\\s+","Platelet", birds_all_years$Subsite)


head(birds_all_years)
```

```{r}
#Converting numeric weather variables####

# Convert surface_water values to numeric
birds_all_years$surface_water <- as.numeric(gsub("%", "", birds_all_years$surface_water)) / 100

# Convert other weather columns to numeric
climate_cols <- c("cloudiness", "rain", "wind", "visibility")
birds_all_years[, climate_cols] <- lapply(birds_all_years[, climate_cols], function(x) as.numeric(as.character(x)))
```

```{r}
#CREATING DATASET FOR SURFACE_WATER TREND ANALYSIS####

# Ensure numeric and factor columns
birds_all_years <- birds_all_years %>%
  mutate(
    Date = as.Date(Date, format="%m/%d/%Y"),
    Site = as.factor(Site),
    `1er ou 2e passage` = as.factor(`1er ou 2e passage`)
  )
birds_all_years <- birds_all_years %>%
  mutate(
    year = format(Date, "%Y"),
    .after = Date
  ) %>%
  relocate(site = Site, passage = `1er ou 2e passage`, .after = year)

birds_all_years <- birds_all_years %>%
  filter(!is.na(year) & !is.na(Subsite))

```

```{r}
library(tidyr)
#Extracting Species Data into long format####

# Metadata columns
meta_names <- 1:16

# Species columns
species_cols <- (17:ncol(birds_all_years))

birds_long <- birds_all_years %>%
  pivot_longer(
    cols = all_of(species_cols),
    names_to = "species",
    values_to = "abundance"
  )
birds_long <- birds_long %>%
  mutate(
    abundance = trimws(abundance),              # remove spaces
    abundance = ifelse(abundance == "", NA, abundance),
    abundance = as.numeric(abundance)           # convert to numeric
  )

species_tobe_removed <- birds_long %>%
  group_by(species) %>%
  summarise(total_abundance = sum(abundance, na.rm = TRUE)) %>%
  filter(total_abundance == 0) %>%
  pull(species)

birds_long <- birds_long %>%
  filter(!species %in% species_tobe_removed)  

```

```{r}
#REMOVING NON-BREEDING SPECIES####
species_info <- read_excel("C:/Users/Lenovo/Documents/BTR/Bird_Data/Espece nicheur-revMT.xlsx")

species_info <- species_info %>%
  rename(
    species = Noms.Especes,
    migrant = `Migrant/Vagrant`
  )

birds_filtered <- birds_long %>%
  left_join(species_info, by = "species")

birds_filtered <- birds_filtered %>%
  filter(!grepl("^1", migrant) | is.na(migrant))

birds_filtered <- birds_filtered %>%
  mutate(line_id = paste(Subsite, Date, passage, sep = "_"))

line_abundance <- birds_filtered %>%
  group_by(line_id) %>%
  summarise(
    abundance = sum(abundance, na.rm = TRUE),
    .groups = "drop"
  )

line_metadata <- birds_filtered %>%
  select(line_id, Subsite, `Type (Restauration active / passive / control)`, julian_day, Date, year, site, passage,
         `Heure debut`, cloudiness, rain, wind, visibility, surface_water, `Observer Name`, Department) %>%
  distinct(line_id, .keep_all = TRUE)

glmm_df <- line_metadata %>%
  left_join(line_abundance, by = "line_id")

glmm_df <- glmm_df %>%
  filter(!(line_id == "Briscon10_2024_1" & surface_water == 0.01),
         !(line_id == "PourtourRascaillan10_2019_2" & surface_water == 1),
         !(line_id == "PourtourRascaillan4_2017_1" & wind == 1))

#CREATING GLMM####
glmm_df
```

```{r}
length(unique(birds_filtered$species))

subset(birds_filtered,birds_filtered$species=="Grand.duc.d.Europe"&birds_filtered$abundance>1)
sort(tapply(birds_filtered$abundance,birds_filtered$species,sum,na.rm=T))

birds_filtered[birds_filtered$Subsite=="Briscon4"&birds_filtered$year=="2018" , ]
```

```{r}
#There are four more rows in the other dataset, but Pourtrague has 4 extra (incorrect) observations for 2018! Which means 1599 is the correct number of lines!!!!!
species_by_line <- birds_filtered %>%
  group_by(line_id, species) %>%
  summarise(
    abundance = sum(abundance, na.rm = TRUE),
    .groups = "drop"
  )

community_matrix <- species_by_line %>%
  pivot_wider(
    names_from = species,
    values_from = abundance,
    values_fill = 0
  )

metadata <- birds_filtered %>%
  select(
    line_id,
    Subsite,
    `Type (Restauration active / passive / control)`,
    julian_day,
    Date,
    year,
    site,
    passage,
    `Heure debut`,
    cloudiness,
    rain,
    wind,
    visibility,
    surface_water,
    `Observer Name`,
    Department
  ) %>%
  distinct(line_id, .keep_all = TRUE)

community_matrix_subsite <- metadata %>%
  left_join(community_matrix, by = "line_id")


community_matrix_subsite
```
```{r}
length(unique(birds_filtered$species))
```


```{r}
species_mat <- as.matrix(community_matrix_subsite[, -(1:16)])

alpha_df <- data.frame(
  richness = specnumber(species_mat),
  shannon  = diversity(species_mat, "shannon")
)

alpha_df$evenness <- ifelse(
  alpha_df$richness > 1,
  alpha_df$shannon / log(alpha_df$richness),
  NA
)

community_matrix_subsite <- bind_cols(
  community_matrix_subsite,
  alpha_df
)


glmm_df <- glmm_df %>%
  left_join(
    community_matrix_subsite %>%
      select(line_id, richness, shannon, evenness),
    by = "line_id"
  )

```
```{r}
library(stringr)

glmm_df <- glmm_df %>%
  mutate(
    Subsite = str_trim(Subsite),                 # remove leading/trailing spaces
    Subsite = str_replace_all(Subsite, "\\s+", ""),  # remove ALL spaces
  Subsite = str_replace(Subsite, "^Enfores", "Enfor"),  # enfores → enfor
    Subsite = str_replace(Subsite, "^Enfor(?=\\d)", "Enfor") # keep format clean
    )
```

```{r}
coordinates <- read.csv("C:/Users/Lenovo/Documents/BTR/Bird_Data/Coordonnees Points STOC-EPS3_Lambert2étendu.csv")
names(coordinates)[c(1, 2, 3)] <- c("Subsite", "x_coord", "y_coord")

#Converting to latitude and longitude values####
library(sf)
coords_sf <- st_as_sf(
  coordinates,
  coords = c("x_coord", "y_coord"),
  crs = 32631   # UTM WGS84 zone 31N
)

coords_latlon <- st_transform(coords_sf, crs = 4326)

coords_latlon <- coords_latlon %>%
  mutate(
    lon = st_coordinates(.)[,1],
    lat = st_coordinates(.)[,2]
  )

glmm_df <- merge(
  glmm_df,
  coords_latlon[, c("Subsite",
                  "lon",
                  "lat",
                  "X..Lambert.II.étendu.",
                  "Y..Lambert.II.étendu.")],
  by = "Subsite",
  all.x = TRUE
)

birds_2024 <- birds_2024[-1, ]
names(birds_2024)[1] <- "Subsite"
names(birds_2024)[3] <- "Type de Restauration"


birds_2024 <- birds_2024 %>%
  mutate(
    `Type de Restauration` = case_when(
      `Type de Restauration` %in% c("Témoin (Bélugue, DPM)") ~ "Control",
      `Type de Restauration` == "Anciens salins - passif" ~ "Passive",
      `Type de Restauration` == "Anciens salins - actif" ~ "Active",
      `Type de Restauration` == "anciens salins - actif" ~ "Active",
      TRUE ~ `Type de Restauration`  # keep everything else unchanged
    )
  )
control <- birds_2024%>%
  select(Subsite, `Type de Restauration`)

control_unique <- control %>%
  select(Subsite, `Type de Restauration`) %>%
  distinct()

glmm_df  <- left_join(glmm_df, control_unique, by = "Subsite")

glmm_df <- glmm_df %>%
  rename(restauration=`Type de Restauration` )

glmm_df <- glmm_df %>%
  mutate(
    across(c(cloudiness, wind, rain, visibility),
           ~ ifelse(. == 0, 1, .))
  )

glmm_df$cloudiness <- as.factor(as.character(glmm_df$cloudiness))
glmm_df$wind <- as.factor(as.character(glmm_df$wind))
```

```{r}
#Vegetation cover!!!#####
vegetation_cover <- read.csv("C:/Users/Lenovo/Documents/BTR/Statistics/vegetation_cover_corr.csv")
names(vegetation_cover)[2] <- "initial_cover"
names(vegetation_cover)[3] <- "final_cover"
names(vegetation_cover)[4] <- "vegetation_change"
setdiff(glmm_df$Subsite, vegetation_cover$Subsite)
setdiff(vegetation_cover$subsite, glmm_df$subsite)
length(intersect(glmm_df$subsite, vegetation_cover$subsite))

vegetation_cover$initial_cover <- as.numeric(gsub("%", "", vegetation_cover$initial_cover))
vegetation_cover$final_cover <- as.numeric(gsub("%", "", vegetation_cover$final_cover))
vegetation_cover$vegetation_change <- as.numeric(gsub("%", "", vegetation_cover$vegetation_change))

library(dplyr)

glmm_df <- glmm_df %>%
  left_join(vegetation_cover, by = "Subsite")

change_cover_test <- lm(vegetation_change ~ restauration, data = glmm_df)

summary(change_cover_test)
emmeans(change_cover_test, pairwise ~ restauration)

emmeans(change_cover_test, pairwise ~ restauration)

plot(allEffects(change_cover_test), type="response")


glmm_df$success <- glmm_df$initial_cover
glmm_df$failure <- 100 - glmm_df$initial_cover

model <- glm(
  cbind(success, failure) ~ restauration,
  data = glmm_df,
  family = binomial
)
plot(allEffects(model), type="response")

summary(model)
```
```{r}
emmeans(model, ~ restauration, type = "response")
```


```{r}
library(ggeffects)
eff_lm <- ggpredict(change_cover_test, terms = "restauration")

ggplot(eff_lm, aes(x = x, y = predicted, group = 1)) +
  geom_point(size = 3) +
    geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.1) +
  labs(
    x = "Restoration type",
    y = "Vegetation change (%)"
  ) +
  theme_minimal(base_size = 14)

eff_glm <- ggpredict(model, terms = "restauration")

ggplot(eff_glm, aes(x = x, y = predicted * 100, group = 1)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = conf.low * 100, ymax = conf.high * 100), width = 0.1) +
  labs(
    x = "Restoration type",
    y = "Initial vegetation cover (%)"
  ) +
  theme_minimal(base_size = 14)
```



```{r}
glmm_df$lon_red <- round(glmm_df$lon,4)
glmm_df$lat_red <- round(glmm_df$lat,4)
glmm_df$pos <- numFactor(glmm_df$lon_red, glmm_df$lat_red) #create numerical factor to create geographic position
glmm_df$group <- factor(rep(1, nrow(glmm_df))) #chercher la justification sur la page glmm_TMB spatial correlation

glmm_df$julian_day <- as.numeric(as.character(glmm_df$julian_day))


#GLMM Richness####
glmm_df$year_num <- as.numeric(as.character(glmm_df$year))

glmm_df <- glmm_df %>%
  filter(
    !is.na(richness),
    !is.na(lat),
    !is.na(lon)
  )
#var(glmm_df$richness)
#mean(glmm_df$richness)
glmm_df$log_water <- log(glmm_df$surface_water+1)

#as numeric
mod_richness_num <- glmmTMB(
  richness ~ scale(log(surface_water+1)) + scale(year_num)*restauration+cloudiness+wind+scale(julian_day)+ scale(initial_cover)*scale(vegetation_change)+(1|site/Subsite), #+ exp(pos + 0 | group),
  family = poisson(),
  data = glmm_df
)
#####################################################

#as factor
mod_richness_fac <- glmmTMB(
  richness ~ scale(log(surface_water+1)) + year*restauration + cloudiness+wind+scale(julian_day)+scale(initial_cover)*scale(vegetation_change)+(1|site/Subsite), #+ exp(pos + 0 | group),
  family = poisson(),
  data = glmm_df
)
plot(allEffects(mod_richness_fac), type="response")
summary(mod_richness_num)
car::Anova(mod_richness_num)

# numeric year model
df_richness_num <- as.data.frame(allEffects(mod_richness_num)[[5]])
df_richness_num$model <- "numeric"

# factor year model
df_richness_fac <- as.data.frame(allEffects(mod_richness_fac)[[5]])
df_richness_fac$model <- "factor"
df_richness_fac$year_num <- as.numeric(as.character(df_richness_fac$year))

ggplot() +
  # --- NUMERIC MODEL (smooth trend) ---
  geom_line(data = df_richness_num,
            aes(x = year_num, y = fit, colour = restauration),
            size = 1) +

  geom_ribbon(data = df_richness_num,
              aes(x = year_num, ymin = lower, ymax = upper, fill = restauration),
              alpha = 0.2) +

  # --- FACTOR MODEL (year-specific points) ---
  geom_point(data = df_richness_fac,
             aes(x = year_num, y = fit, colour = restauration),
             size = 2) +
  geom_area() + scale_x_continuous(breaks=min(glmm_df$year):max(glmm_df$year))+
  labs(x = "Year",
       y = "Predicted Species Richness",
       colour = "Restoration",
       fill = "Restoration") +
  theme_classic()

simulationOutput <- simulateResiduals(fittedModel = mod_richness_num, plot = T)
```

```{r}
#GLMM Abundance####

#numeric
mod_abundance_num <- glmmTMB(
  abundance ~ scale(log(surface_water+1)) + scale(year_num) * restauration+ cloudiness+wind+scale(julian_day) +scale(initial_cover)*scale(vegetation_change) +(1|site/Subsite), #+ exp(pos + 0 | group),
  family = nbinom2(), #the mean, therefore take nbinom2
  data = glmm_df
  )

#factor
mod_abundance_fac <- glmmTMB(
  abundance ~ scale(log(surface_water+1)) + year * restauration+cloudiness+wind+ scale(julian_day)+scale(initial_cover)*scale(vegetation_change) +(1|site/Subsite), #+ exp(pos + 0 | group),
  family = nbinom2(), #the mean, therefore take nbinom2
  data = glmm_df
  )

# numeric year model
df_abundance_num <- as.data.frame(allEffects(mod_abundance_num)[[5]])
df_abundance_num$model <- "numeric"

# factor year model
df_abundance_fac <- as.data.frame(allEffects(mod_abundance_fac)[[5]])
df_abundance_fac$model <- "factor"
df_abundance_fac$year_num <- as.numeric(as.character(df_abundance_fac$year))

#mean(glmm_df$abundance)

hist(glmm_df$abundance)
plot(allEffects(mod_abundance_fac), type="response")
summary(mod_abundance_num)
car::Anova(mod_abundance_num)

ggplot() +
  # --- NUMERIC MODEL (smooth trend) ---
  geom_line(data = df_abundance_num,
            aes(x = year_num, y = fit, colour = restauration),
            size = 1) +

  geom_ribbon(data = df_abundance_num,
              aes(x = year_num, ymin = lower, ymax = upper, fill = restauration),
              alpha = 0.2) +

  # --- FACTOR MODEL (year-specific points) ---
  geom_point(data = df_abundance_fac,
             aes(x = year_num, y = fit, colour = restauration),
             size = 2) +
  geom_area() + scale_x_continuous(breaks=min(glmm_df$year):max(glmm_df$year))+
  
  labs(x = "Year",
       y = "Predicted Total Abundance",
       colour = "Restoration",
       fill = "Restoration") +
  theme_classic()

simulationOutput <- simulateResiduals(fittedModel = mod_abundance_fac, plot = T)
#testDispersion(simulationOutput)
```

```{r}
#https://cran.r-project.org/web/packages/emmeans/vignettes/comparisons.html

#GLMM Eveness####

#as numeric
glmm_df$evenness_cor <- glmm_df$evenness - 0.001
mod_eveness_num <- glmmTMB(
  evenness_cor ~ scale(log(surface_water+1)) + scale(year_num)*restauration+cloudiness+wind+scale(julian_day)+scale(initial_cover)*scale(vegetation_change) +(1|site/Subsite), # + exp(pos + 0 | group),
  family = beta_family(),
  data = glmm_df
)

#as factor
glmm_df$evenness_cor <- glmm_df$evenness - 0.01
mod_eveness_fac <- glmmTMB(
  evenness_cor ~ scale(log(surface_water+1)) + year*restauration +cloudiness+wind+scale(julian_day)+scale(initial_cover)*scale(vegetation_change) +(1|site/Subsite), # + exp(pos + 0 | group),
  family = beta_family(),
  data = glmm_df
)

# numeric year model
df_even_num <- as.data.frame(allEffects(mod_eveness_num)[[5]])
df_even_num$model <- "numeric"

# factor year model
df_even_fac <- as.data.frame(allEffects(mod_eveness_fac)[[5]])
df_even_fac$model <- "factor"
df_even_fac$year_num <- as.numeric(as.character(df_even_fac$year))

plot(allEffects(mod_eveness_fac), type="response")
summary(mod_eveness_num)
car::Anova(mod_eveness_num)

ggplot() +
  # --- NUMERIC MODEL (smooth trend) ---
  geom_line(data = df_even_num,
            aes(x = year_num, y = fit, colour = restauration),
            size = 1) +

  geom_ribbon(data = df_even_num,
              aes(x = year_num, ymin = lower, ymax = upper, fill = restauration),
              alpha = 0.2) +

  # --- FACTOR MODEL (year-specific points) ---
  geom_point(data = df_even_fac,
             aes(x = year_num, y = fit, colour = restauration),
             size = 2) +
  geom_area() + scale_x_continuous(breaks=min(glmm_df$year):max(glmm_df$year))+
  
  labs(x = "Year",
       y = "Predicted Eveness",
       colour = "Restoration",
       fill = "Restoration") +
  theme_classic()

simulationOutput <- simulateResiduals(fittedModel = mod_eveness_fac, plot = T)
```


#https://cran.r-project.org/web/packages/emmeans/vignettes/comparisons.html

#GLMM Eveness####
glmm_df$eveness_cor <- glmm_df$eveness - 0.01
mod_eveness <- glmmTMB(
  eveness_cor ~ log(surface_water*100+1) + year_num*restauration +site+(1|Subsite), #+ mat(pos + 0 | group)
  family = beta_family(),
  data = glmm_df
)
# numeric year model
df_num <- as.data.frame(allEffects(mod_abundance_num)[[4]])
df_num$model <- "numeric"

# factor year model
df_fac <- as.data.frame(allEffects(mod_abundance_fac)[[4]])
df_fac$model <- "factor"
df_fac$year_num <- as.numeric(as.character(df_fac$year))

plot(allEffects(mod_eveness), type="response")
summary(mod_eveness)
car::Anova(mod_eveness)

anova_eveness <- as.data.frame(car::Anova(mod_eveness))
anova_eveness$variable <- "eveness"
anova_eveness$Chisq <- round(anova_eveness$Chisq,3)

emtrends(mod_eveness,~ restauration,var=  "year_num")

df <- as.data.frame(allEffects(mod_eveness)[[4]])
df$variable <- "eveness"
  

ggplot(df)+
  # geom_point(aes(x=year_num, y=fit, col=restauration))+
  geom_line(aes(x=year_num, y=fit, col=restauration))+
  geom_ribbon(aes(x=year_num, ymin=lower, ymax=upper, fill=restauration),alpha=0.2)+
labs(x = "Year", y = "Eveness", colour = "Restoration", fill = "Restoration")

simulationOutput <- simulateResiduals(fittedModel = mod_eveness, plot = T)


# year as factor
glmm_df$eveness_cor <- glmm_df$eveness - 0.01
mod_eveness_yrfac <- glmmTMB(
  eveness_cor ~ log(surface_water*100+1) + year*restauration +site+(1|Subsite), #+ mat(pos + 0 | group)
  family = beta_family(),
  data = glmm_df
)
summary(glmm_df$eveness)

plot(allEffects(mod_eveness_yrfac), type="response")
summary(mod_eveness_yrfac)
car::Anova(mod_eveness_yrfac)

anova_eveness_yrfac <- as.data.frame(car::Anova(mod_eveness_yrfac))
anova_eveness_yrfac$variable <- "eveness"
anova_eveness_yrfac$Chisq <- round(anova_eveness$Chisq,3)

(mod_eveness_yrfac,pairwise~year*restauration,type="response", adjust = "bonferroni")

df_yrfac <- as.data.frame(allEffects(mod_eveness_yrfac)[[3]])
df_yrfac$variable <- "eveness"
  

ggplot(df_yrfac)+
  # geom_point(aes(x=year_num, y=fit, col=restauration))+
  geom_line(aes(x=year, y=fit, col=restauration))+
  geom_ribbon(aes(x=year, ymin=lower, ymax=upper, fill=restauration),alpha=0.2)+
labs(x = "Year", y = "Eveness", colour = "Restoration", fill = "Restoration")

```{r}
#Shannon####
#as numeric

mod_shannon_num <- glmmTMB(
  shannon ~ scale(log(surface_water+1))+ scale(year_num)*restauration+cloudiness+wind+ scale(julian_day) + scale(initial_cover)*scale(vegetation_change)  +(1|site/Subsite), # exp(pos + 0 | group),
  family = gaussian(),
  data = glmm_df
)

#as factor
mod_shannon_fac <- glmmTMB(
  shannon ~ scale(log(surface_water+1))+ year*restauration +cloudiness+wind+ scale(julian_day) +scale(initial_cover)*scale(vegetation_change) + (1|site/Subsite), #exp(pos + 0 | group),
  family = gaussian(),
  data = glmm_df
)

# numeric year model
df_shannon_num <- as.data.frame(allEffects(mod_shannon_num)[[5]])
df_shannon_num$model <- "numeric"

# factor year model
df_shannon_fac <- as.data.frame(allEffects(mod_shannon_fac)[[5]])
df_shannon_fac$model <- "factor"
df_shannon_fac$year_num <- as.numeric(as.character(df_shannon_fac$year))

plot(allEffects(mod_shannon_fac), type="response")
summary(mod_shannon_num)
car::Anova(mod_shannon_num)


ggplot() +
  # --- NUMERIC MODEL (smooth trend) ---
  geom_line(data = df_shannon_num,
            aes(x = year_num, y = fit, colour = restauration),
            size = 1) +

  geom_ribbon(data = df_shannon_num,
              aes(x = year_num, ymin = lower, ymax = upper, fill = restauration),
              alpha = 0.2) +

  # --- FACTOR MODEL (year-specific points) ---
  geom_point(data = df_shannon_fac,
             aes(x = year_num, y = fit, colour = restauration),
             size = 2) +
  geom_area() + scale_x_continuous(breaks=min(glmm_df$year):max(glmm_df$year))+
  
  labs(x = "Year",
       y = "Predicted Shannon",
       colour = "Restoration",
       fill = "Restoration") +
  theme_classic()


simulationOutput <- simulateResiduals(fittedModel = mod_shannon_fac, plot = T)
```

```{r}
#COMPARE AIC FOR MODEL VALIDATION - REDUNDANT ####
m1 <- glmmTMB(
  shannon ~ log(surface_water+1) + year_num*restauration+ site + (1|Subsite), #best for abundance + best for evenness (but is negative) + best for shannon
  family = gaussian(),
  data = glmm_df
)

m2 <- glmmTMB(
  shannon ~ log(surface_water+1) + year_num *restauration+site+passage +(1|Subsite),  #best for richness
  family = gaussian(),
  data = glmm_df
)

m3 <- glmmTMB(
  shannon ~ log(surface_water+1)+year_num*restauration+passage+(1|Subsite), 
  family = gaussian(),
  data = glmm_df
)

AIC(m1, m2, m3)
```
```{r}
#TEST FOR SPATIAL-CORRELATION FOR 4 VARIABLES####

models <- list(
  abundance = mod_abundance_fac,
  richness = mod_richness_fac,
  eveness  = mod_eveness_fac,
  shannon  = mod_shannon_fac
)

temporaire <- glmm_df[!duplicated(glmm_df$Subsite), ]

results <- list()

for(name in names(models)) {
  
  cat("### Testing spatial autocorrelation for", name, "###\n")
  
  sim_out <- simulateResiduals(models[[name]])
  
  sim_grouped <- recalculateResiduals(sim_out, group = temporaire$Subsite)
  
  test_result <- testSpatialAutocorrelation(sim_grouped, x = temporaire$lon, y = temporaire$lat, plot = TRUE)
  
  results[[name]] <- test_result
  
  print(test_result)
}
```
```{r}
# isolate ONLY species columns ONCE
species_mat <- as.matrix(community_matrix_subsite[, -(c(1:16, (ncol(community_matrix_subsite)-2):ncol(community_matrix_subsite)))])

sac <- specaccum(species_mat, method = "random")

plot(sac, ci.type = "polygon", ci.col = "lightblue", col = "darkblue", lwd= 2, xlab = "Sampling Effort", ylab="Species Richness")
```
```{r}
sac <- specaccum(species_mat, method = "random")

plot(sac,
     ci.type = "polygon",
     ci.col = "lightblue",
     col = "darkblue",
     lwd = 2,
     xlab = "Sampling Effort",
     ylab = "Species Richness")

# interpolate richness at exactly 100 surveys
richness_100 <- approx(sac$sites, sac$richness, xout = 100)$y

# percentage of total richness
percent_100 <- (richness_100 / 96) * 100

richness_100
percent_100
```

```{r}
community_matrix_subsite <- community_matrix_subsite %>%
  mutate(site = str_replace_all(site, "Pourtour\\s*Rascaillan|PourtourRascaillan", "Pourtour"))

#Site SAC with sampling effort = Subsites####

# Get unique sites
sites <- unique(community_matrix_subsite$site)

# List to store SACs
sac_list <- list()

# Loop over each site
for (s in sites) {
  # Select subsites for this site
  subsites_mat <- species_mat[community_matrix_subsite$site == s, ]
  
  # Make sure it's still a matrix if only one row
  if (is.vector(subsites_mat)) {
    subsites_mat <- matrix(subsites_mat, nrow = 1)
  }
  
  # Species accumulation
  sac_list[[s]] <- specaccum(subsites_mat, method = "random")
}

par(mfrow = c(2, 5)) # 2 rows, 5 columns if you have 10 sites

for (s in sites) {
  plot(sac_list[[s]], main = paste("Site", s), xlab = "Subsites", ylab = "Species")
}
```
```{r}

# List to store SACs
sites <- unique(community_matrix_subsite$site)
sac_year_list <- list()

for (s in sites) {
  # Select rows for this site
  site_data <- community_matrix_subsite[community_matrix_subsite$site == s, ]
  
  # Extract only species columns (metadata = first 16 columns)
  species_data <- site_data[, -(c(1:16, (ncol(site_data)-2):ncol(site_data)))]
  
  # Make sure species columns are numeric
  species_data[] <- lapply(species_data, as.numeric)
  
  # Aggregate subsites by year
  site_year_mat <- aggregate(species_data, by = list(year = site_data$year), FUN = sum)
  
  # Convert to matrix (rows = years, cols = species)
  species_mat_year <- as.matrix(site_year_mat[, -1])  # drop the 'year' column
  
  # SAC over years (chronological order)
  sac_year_list[[s]] <- specaccum(species_mat_year, method = "collector")
}

par(mfrow = c(2,5))
for (s in sites) {
  plot(sac_year_list[[s]], main = paste("Site", s),
       xlab = "Year samples", ylab = "Species richness")
}
```
```{r}
# Get unique sites
sites <- unique(community_matrix_subsite$site)

for (s in sites) {
  # Select data for this site
  site_data <- community_matrix_subsite[community_matrix_subsite$site == s, ]
  
  # Remove rows with missing or empty passage
  site_data <- site_data[!is.na(site_data$passage) & site_data$passage != "", ]
  
  # Skip this site if no valid rows
  if (nrow(site_data) == 0) next
  
  # Determine passages present for this site
  site_passages <- unique(site_data$passage)
  
  # List to store SACs for each passage
  sac_passage_list <- list()
  
  for (p in site_passages) {
    # Select data for this passage
    passage_data <- site_data[site_data$passage == p, ]
    
    # Extract species columns (columns 16 onward)
    species_data <- passage_data[, -(c(1:16, (ncol(passage_data)-2):ncol(passage_data)))]
    species_data[] <- lapply(species_data, as.numeric)
    
    # Convert to matrix
    species_mat <- as.matrix(species_data)
    
    # Only calculate SAC if at least one row exists
    if (nrow(species_mat) > 0) {
      sac_passage_list[[as.character(p)]] <- specaccum(species_mat, method = "collector")
    }
  }
  
  # Skip plotting if no valid passages
  if (length(sac_passage_list) == 0) next
  
  # Assign colors: handle 1–2 passages manually
  n_passages <- length(sac_passage_list)
  if (n_passages == 1) {
    colors <- "red"
  } else if (n_passages == 2) {
    colors <- c("red","blue")
  } else {
    colors <- RColorBrewer::brewer.pal(n_passages, "Set1")
  }
  
  # Plot first passage
  # Plot first passage
plot(sac_passage_list[[1]]$sites, sac_passage_list[[1]]$richness, type = "l",
     col = colors[1], lwd = 1,  # thin line
     main = paste("Site", s),
     xlab = "Sampling effort", ylab = "Species richness",
     ylim = c(0, max(sapply(sac_passage_list, function(x) max(x$richness)))))


# Overlay additional passages
if (n_passages > 1) {
  for (i in 2:n_passages) {
    lines(sac_passage_list[[i]]$sites, sac_passage_list[[i]]$richness,
          col = colors[i], lwd = 0.1)  # same thin line
    
  }
}

# Legend: match line width to plot
legend("bottomright", legend = names(sac_passage_list), col = colors,
       lwd = 1, pch = 16, title = "Passage")
}
```
```{r}
#YEAR*RESTORATION INTERACTION####
df_shannon_num$metric <- "Shannon"
df_richness_num$metric <- "Richness"
df_even_num$metric <- "Evenness"
df_abundance_num$metric <- "Abundance"

df_num_all <- rbind(df_shannon_num, df_richness_num, df_even_num, df_abundance_num)


df_shannon_fac$metric <- "Shannon"
df_richness_fac$metric <- "Richness"
df_even_fac$metric <- "Evenness"
df_abundance_fac$metric <- "Abundance"

df_fac_all <- rbind(df_shannon_fac, df_richness_fac, df_even_fac, df_abundance_fac)
```

```{r}
ggplot() +
  # --- NUMERIC MODEL (smooth trend) ---
  geom_line(data = df_num_all,
            aes(x = year_num, y = fit, colour = restauration),
            linewidth = 1) +

  geom_ribbon(data = df_num_all,
              aes(x = year_num, ymin = lower, ymax = upper, fill = restauration),
              alpha = 0.2) +

  # --- FACTOR MODEL (year-specific points) ---
  geom_point(data = df_fac_all,
             aes(x = year_num, y = fit, colour = restauration),
             size = 2) +
  
  geom_errorbar(data = df_fac_all,
              aes(x = year_num, ymin = lower, ymax = upper, colour = restauration),
              width = 0.2)+
  
  scale_x_continuous(breaks = min(glmm_df$year):max(glmm_df$year)) +

  facet_wrap(~ metric, scales = "free_y") +

  labs(x = "Year",
       y = "Predicted value",
       colour = "Restoration",
       fill = "Restoration") +

  theme_classic()
```
```{r}
#SURFACE WATER####
# --- Prepare dataframes for each index ---
df_shannon <- as.data.frame(allEffects(mod_shannon_num)[[1]])
df_shannon$metric <- "Shannon"

df_evenness <- as.data.frame(allEffects(mod_eveness_num)[[1]])
df_evenness$metric <- "Evenness"

df_richness <- as.data.frame(allEffects(mod_richness_num)[[1]])
df_richness$metric <- "Richness"

df_abundance <- as.data.frame(allEffects(mod_abundance_num)[[1]])
df_abundance$metric <- "Abundance"

# --- Combine into one long dataframe ---
df_surface_all <- rbind(df_shannon, df_evenness, df_richness, df_abundance)

# ---- Significance stars ----
get_p_surface <- function(model) {
  a <- car::Anova(model)
  a$`Pr(>Chisq)`[grep("surface_water", rownames(a))]
}
p_shannon   <- get_p_surface(mod_shannon_num)
p_evenness  <- get_p_surface(mod_eveness_num)
p_richness  <- get_p_surface(mod_richness_num)
p_abundance <- get_p_surface(mod_abundance_num)

get_stars <- function(p) {
  if (is.na(p)) return("")
  if (p < 0.001) "***"
  else if (p < 0.01) "**"
  else if (p < 0.05) "*"
  else ""
}
stars_df <- data.frame(
  metric = c("Shannon", "Evenness", "Richness", "Abundance"),
  stars = c(
    get_stars(p_shannon),
    get_stars(p_evenness),
    get_stars(p_richness),
    get_stars(p_abundance)
  )
)
stars_df <- data.frame(
  metric = c("Shannon", "Evenness", "Richness", "Abundance"),
  stars = c(
    get_stars(p_shannon),
    get_stars(p_evenness),
    get_stars(p_richness),
    get_stars(p_abundance)
  )
)

stars_df <- stars_df %>%
  left_join(
    df_surface_all %>%
      group_by(metric) %>%
      summarise(
        x = max(log(surface_water + 1), na.rm = TRUE),
        y = max(fit, na.rm = TRUE)
      ),
    by = "metric"
  )


# --- Plot ---
ggplot(df_surface_all, aes(x = log(surface_water + 1), y = fit)) +
  geom_line(linewidth = 1) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
  
  geom_text(
    data = stars_df,
    aes(label = stars),
    x = Inf,
    y = Inf,
    hjust = 6,
    vjust = 1.7,
    size = 6,
    inherit.aes = FALSE
  ) +
  
  facet_wrap(~ metric, scales = "free_y") +
  
  scale_x_continuous(
    name = "Surface Water (%)",
    labels = function(x) round(x * 100, 1)
  ) +
  
  theme_classic()

```


```{r}
#CREATING ANOVA DATA FRAMES####

#Eveness#
anova_eveness <- as.data.frame(car::Anova(mod_eveness_num))
anova_eveness$variable <- "eveness"
anova_eveness$Chisq <- round(anova_eveness$Chisq,3)
anova_eveness$`Pr(>Chisq)` <- round(anova_eveness$`Pr(>Chisq)`,3)

#Shannon#
anova_shannon <- as.data.frame(car::Anova(mod_shannon_num))
anova_shannon$variable <- "shannon"
anova_shannon$Chisq <- round(anova_shannon$Chisq,3)
anova_shannon$`Pr(>Chisq)` <- round(anova_shannon$`Pr(>Chisq)`,3)

#Richness#
anova_richness <- as.data.frame(car::Anova(mod_richness_num))
anova_richness$variable <- "richness"
anova_richness$Chisq <- round(anova_richness$Chisq,3)
anova_richness$`Pr(>Chisq)` <- round(anova_richness$`Pr(>Chisq)`,3)

#Abundance#
anova_abundance <- as.data.frame(car::Anova(mod_abundance_num))
anova_abundance$variable <- "abundance"
anova_abundance$Chisq <- round(anova_abundance$Chisq,3)
anova_abundance$`Pr(>Chisq)` <- round(anova_abundance$`Pr(>Chisq)`,3)

#Fix Predictor Column####
anova_shannon$Predictor <- rownames(anova_shannon)
rownames(anova_shannon) <- NULL

anova_eveness$Predictor <- rownames(anova_eveness)
rownames(anova_eveness) <- NULL

anova_richness$Predictor <- rownames(anova_richness)
rownames(anova_richness) <- NULL

anova_abundance$Predictor <- rownames(anova_abundance)
rownames(anova_abundance) <- NULL

#Combine with rbind#
anova_all <- rbind(
  anova_eveness,
  anova_shannon,
  anova_richness,
  anova_abundance
)

```

```{r}
anova_clean <- anova_all[, c("Chisq","Df","Pr(>Chisq)","variable","Predictor")]
colnames(anova_clean) <- c("Chi-square","Df","p-value", "Index", "Predictor")

format_p <- function(p) {
  ifelse(p < 0.001, "<0.001", round(p, 3))
}
anova_clean$`p-value` <- format_p(anova_clean$`p-value`)

anova_shannon_clean   <- subset(anova_clean, Index == "shannon")
anova_eveness_clean   <- subset(anova_clean, Index == "eveness")
anova_richness_clean  <- subset(anova_clean, Index == "richness")
anova_abundance_clean <- subset(anova_clean, Index == "abundance")

wb <- createWorkbook()

addWorksheet(wb, "Shannon")
writeData(wb, "Shannon", anova_shannon_clean)

addWorksheet(wb, "Evenness")
writeData(wb, "Evenness", anova_eveness_clean)

addWorksheet(wb, "Richness")
writeData(wb, "Richness", anova_richness_clean)

addWorksheet(wb, "Abundance")
writeData(wb, "Abundance", anova_abundance_clean)

saveWorkbook(wb, "ANOVA_results.xlsx", overwrite = TRUE)
```

```{r}
#EMMEANS FOR POST_HOC####
emtrends(mod_eveness_num,~ restauration,var=  "year_num")
```
```{r}
emm_eveness <- emmeans(mod_eveness_fac, ~ year * restauration, type = "response")
pairs_eveness <- pairs(emm_eveness, adjust = "bonferroni")
pairs_eveness

emm_shannon <- emmeans(mod_shannon_fac, ~ year*restauration, type = "response")
pairs_shannon <- pairs(emm_shannon, adjust = "bonferroni")
pairs_shannon

emm_richness <- emmeans(mod_richness_fac, ~ year * restauration, type = "response")
pairs_richness <- pairs(emm_richness, adjust = "bonferroni")
pairs_richness

emm_abundance <- emmeans(mod_abundance_fac, ~ year * restauration, type = "response")
pairs_abundance <- pairs(emm_abundance, adjust = "bonferroni")
pairs_abundance
```

```{r}
#TESTING CLIMATE VARIABLES --> COVARIANCE####
cor.test(glmm_df$cloudiness, glmm_df$visibility) #correlated
cor.test(glmm_df$visibility, glmm_df$rain) # not correlated
cor.test(glmm_df$surface_water, glmm_df$rain) #correlated
cor.test(glmm_df$wind, glmm_df$cloudiness) #not correlated
cor.test(glmm_df$wind, glmm_df$visibility) #not correlated
```
```{r}
#SURFACE WATER####
# --- Prepare dataframes for each index ---
df_shannon <- as.data.frame(allEffects(mod_shannon_num)[[2]])
df_shannon$metric <- "Shannon"

df_evenness <- as.data.frame(allEffects(mod_eveness_num)[[2]])
df_evenness$metric <- "Evenness"

df_richness <- as.data.frame(allEffects(mod_richness_num)[[2]])
df_richness$metric <- "Richness"

df_abundance <- as.data.frame(allEffects(mod_abundance_num)[[2]])
df_abundance$metric <- "Abundance"

# --- Combine into one long dataframe ---
df_surface_all <- rbind(df_shannon, df_evenness, df_richness, df_abundance)

# ---- Significance stars ----
get_p_surface <- function(model) {
  a <- car::Anova(model)
  a$`Pr(>Chisq)`[grep("cloudiness", rownames(a))]
}
p_shannon   <- get_p_surface(mod_shannon_num)
p_evenness  <- get_p_surface(mod_eveness_num)
p_richness  <- get_p_surface(mod_richness_num)
p_abundance <- get_p_surface(mod_abundance_num)

get_stars <- function(p) {
  if (is.na(p)) return("")
  if (p < 0.001) "***"
  else if (p < 0.01) "**"
  else if (p < 0.05) "*"
  else if (p < 0.1) "."
  else ""
}
stars_df <- data.frame(
  metric = c("Shannon", "Evenness", "Richness", "Abundance"),
  stars = c(
    get_stars(p_shannon),
    get_stars(p_evenness),
    get_stars(p_richness),
    get_stars(p_abundance)
  )
)
stars_df <- data.frame(
  metric = c("Shannon", "Evenness", "Richness", "Abundance"),
  stars = c(
    get_stars(p_shannon),
    get_stars(p_evenness),
    get_stars(p_richness),
    get_stars(p_abundance)
  )
)

stars_df <- stars_df %>%
  left_join(
    df_surface_all %>%
      group_by(metric) %>%
      summarise(
        x = max(as.numeric(as.character(cloudiness))),
        y = max(fit)
      ),
    by = "metric"
  )


# --- Plot ---
library(ggplot2)

ggplot(df_surface_all, aes(x = cloudiness, y = fit, group = 1)) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2) +
  geom_text(
  data = stars_df,
  aes(label = stars),
  x = Inf,
  y = Inf,
  hjust = 6,   
  vjust = 1.7,
  size = 6,
  inherit.aes = FALSE
) +
  facet_wrap(~ metric, scales = "free_y") +
  theme_classic()

```
```{r}
targets <- c(25, 50, 75)

effort_at_richness <- sapply(targets, function(t) {
  sac$sites[min(which(sac$richness >= t))]
})

effort_at_richness
```
```{r}
data.frame(
  richness = targets,
  effort = effort_at_richness
)
```
```{r}
diff_richness <- diff(sac$richness)
diff_effort <- diff(sac$sites)

marginal_gain <- diff_richness / diff_effort

marginal_gain
```

```{r}
target <- 0.8 * max(sac$richness)

sac$sites[min(which(sac$richness >= target))]
```
```{r}
species_cols <- colnames(community_matrix_subsite)[17:(ncol(community_matrix_subsite) - 1)]


# STEP 1: take max across passages (per subsite)
subsite_year <- community_matrix_subsite %>%
  group_by(site, Subsite, year) %>%
  summarise(across(all_of(species_cols), max, na.rm = TRUE),
            .groups = "drop")

# STEP 2: sum across subsites to get site-level
site_year_matrix <- subsite_year %>%
  group_by(site, year) %>%
  summarise(across(all_of(species_cols), sum, na.rm = TRUE),
            .groups = "drop")


site_year_matrix$richness <- rowSums(site_year_matrix[, species_cols] > 0)

site_year_richness <- site_year_matrix %>%
  mutate(richness = rowSums(across(all_of(species_cols)) > 0)) %>%
  select(site, year, richness)

head(site_year_richness)
```
```{r}
site_year_richness %>% 
  filter(site == "Rascaillan")
```
```{r}
site_year_richness %>% 
  filter(site == "Val Agricola")
```
```{r}
#FUNCTIONAL DIVERSITY####
library(FD)
library(ade4)

#import functional trait dataset
df_traits <- read_excel("C:/Users/Lenovo/Documents/BTR/Statistics/community_matrix_subsite_EG.xlsx", sheet = "Traits_storchova") # tableau avec les traits en colonne et espèces en ligne

#Adding niche in camargue or LMCS to data:
species_info <- read_excel("C:/Users/Lenovo/Documents/BTR/Bird_Data/Espece nicheur-revMT.xlsx")

species_info <- species_info %>%
  rename(
    Species = Noms.Especes,
    migrant = `Migrant/Vagrant`,
    niche_lmcs = `Niche LMCS`,
    niche_camargue = `Niche Camargue`
  )



df_traits <- df_traits %>%
  left_join(species_info, by = "Species")

#pick variables
df_traits_clean <- as.data.frame(df_traits[,c("Wetland","Grassland","Shrubland","Forest", "Folivore_B", "Granivore_B","Other_invertebrates_B","Fish_B","Incubation_period","niche_camargue", "niche_lmcs")])
rownames(df_traits_clean) <- df_traits$Species

#determine number of axes
traits.dist<-gowdis(df_traits_clean)  # utilisation de l'indice de gower pour faire la matrice de distance
traits.pcoa=cailliez((traits.dist))
pcoa_traits<-dudi.pco(traits.pcoa) # voir combien il faut garder d'axes (ici 4) 
4

#create PCoA
PCOA1<-unlist(pcoa_traits$li[,1])
PCOA2<-unlist(pcoa_traits$li[,2])
PCOA3<-unlist(pcoa_traits$li[,3])
PCOA4<-unlist(pcoa_traits$li[,4])

traits_matrix<-cbind(PCOA1,PCOA2,PCOA3,PCOA4)

rownames(traits_matrix) <- df_traits$Species


#rownames(traits_matrix) <- rownames(df_traits_clean)

#combine with abundance data
data_abondance <- read_excel(
  "C:/Users/Lenovo/Documents/BTR/Statistics/community_matrix_subsite_EG.xlsx",
  sheet = "Data"
) 

#create the same species matrix with the appropriate amount of species, here called data_abondance
data_abondance <- data_abondance[, -(c(1:16, ncol(data_abondance)))]# tableau avec les espèces en colonne
data_abondance <- data_abondance[, rownames(traits_matrix)]

data_abondance<-as.matrix(data_abondance)
row_sums <- rowSums(data_abondance)
data_abondance <- data_abondance[row_sums > 0, ] #remove rows with 0 observations

FDall<-dbFD(traits_matrix,data_abondance)

FRichness<-FDall$FRic
FEvenness<-FDall$FEve
FDivergence<-FDall$FDiv
FSpecialization<-FDall$FSpe
RaoQ <- FDall$RaoQ

FDall
```


```{r}
fd_df <- glmm_df %>%
  mutate(
    FRic = FDall$FRic,
    FEve = FDall$FEve,
    FDiv = FDall$FDiv,
    FSpe = FDall$FSpe,
    RaoQ = FDall$RaoQ
  )

fd_df <- fd_df %>%
  mutate(
    year_num = as.numeric(as.character(year)),
    log_water = log(surface_water + 1)
  )

#FRIC GLMM####

#abundance ~ log(surface_water+1) + year_num * restauration +
#cloudiness + wind + julian_day + (1|site/Subsite)

#fd_df <- fd_df %>%
#  filter(!is.na(FRic))

#numeric
mod_FRic_num <- glmmTMB(
  FRic ~ log(surface_water + 1) + year_num * restauration +
    cloudiness + wind + julian_day,# +(1|site/Subsite),
  family = gaussian(),
  data = fd_df
)

#factor
mod_FRic_fac <- glmmTMB(
  FRic ~ log(surface_water + 1) + year * restauration +
    cloudiness + wind + julian_day,# +(1|site/Subsite),
  family = gaussian(),
  data = fd_df
)

plot(allEffects(mod_FRic_fac), type="response")
summary(mod_FRic_num)
car::Anova(mod_FRic_num)


# numeric year model
df_FRic_num <- as.data.frame(allEffects(mod_FRic_num)[[5]])
df_FRic_num$model <- "numeric"

# factor year model
df_FRic_fac <- as.data.frame(allEffects(mod_FRic_fac)[[5]])
df_FRic_fac$model <- "factor"
df_FRic_fac$year_num <- as.numeric(as.character(df_FRic_fac$year))



ggplot() +
  # --- NUMERIC MODEL (smooth trend) ---
  geom_line(data = df_FRic_num,
            aes(x = year_num, y = fit, colour = restauration),
            size = 1) +

  geom_ribbon(data = df_FRic_num,
              aes(x = year_num, ymin = lower, ymax = upper, fill = restauration),
              alpha = 0.2) +

  # --- FACTOR MODEL (year-specific points) ---
  geom_point(data = df_FRic_fac,
             aes(x = year_num, y = fit, colour = restauration),
             size = 2) +
  geom_area() + scale_x_continuous(breaks=min(glmm_df$year):max(glmm_df$year))+
  geom_errorbar(data = df_FRic_fac,
              aes(x = year_num, ymin = lower, ymax = upper, colour = restauration),
              width = 0.2)+
  
  labs(x = "Year",
       y = "Predicted FRic",
       colour = "Restoration",
       fill = "Restoration") +
  theme_classic()

simulationOutput <- simulateResiduals(mod_FRic_num, plot = TRUE)
```

```{r}
#FEVENNESS####

#numeric
mod_FEvenness_num <- glmmTMB(
  FEvenness ~ log(surface_water + 1) + year_num * restauration +
    cloudiness + wind + julian_day,# +(1|site/Subsite),
  family = gaussian(),
  data = fd_df
)

#factor
mod_FEvenness_fac <- glmmTMB(
  FEvenness ~ log(surface_water + 1) + year * restauration +
    cloudiness + wind + julian_day,# +(1|site/Subsite),
  family = gaussian(),
  data = fd_df
)

plot(allEffects(mod_FEvenness_fac), type="response")
summary(mod_FEvenness_num)
car::Anova(mod_FEvenness_num)


# numeric year model
df_FEvenness_num <- as.data.frame(allEffects(mod_FEvenness_num)[[5]])
df_FEvenness_num$model <- "numeric"

# factor year model
df_FEvenness_fac <- as.data.frame(allEffects(mod_FEvenness_fac)[[5]])
df_FEvenness_fac$model <- "factor"
df_FEvenness_fac$year_num <- as.numeric(as.character(df_FEvenness_fac$year))



ggplot() +
  # --- NUMERIC MODEL (smooth trend) ---
  geom_line(data = df_FEvenness_num,
            aes(x = year_num, y = fit, colour = restauration),
            size = 1) +

  geom_ribbon(data = df_FEvenness_num,
              aes(x = year_num, ymin = lower, ymax = upper, fill = restauration),
              alpha = 0.2) +

  # --- FACTOR MODEL (year-specific points) ---
  geom_point(data = df_FEvenness_fac,
             aes(x = year_num, y = fit, colour = restauration),
             size = 2) +
  geom_area() + scale_x_continuous(breaks=min(glmm_df$year):max(glmm_df$year))+
  geom_errorbar(data = df_FEvenness_fac,
              aes(x = year_num, ymin = lower, ymax = upper, colour = restauration),
              width = 0.2)+
  
  labs(x = "Year",
       y = "Predicted FEvenness",
       colour = "Restoration",
       fill = "Restoration") +
  theme_classic()

simulationOutput <- simulateResiduals(mod_FEvenness_num, plot = TRUE)

```
```{r}
#FDivergence GLMM####

#CHECK JULIAN DAY AS NUMERIC

#numeric
mod_FDivergence_num <- glmmTMB(
  FDivergence ~ log(surface_water + 1) + year_num * restauration +
    cloudiness + wind + julian_day,# +(1|site/Subsite),
  family = gaussian(),
  data = fd_df
)

#factor
mod_FDivergence_fac <- glmmTMB(
  FDivergence ~ log(surface_water + 1) + year * restauration +
    cloudiness + wind + julian_day,# +(1|site/Subsite),
  family = gaussian(),
  data = fd_df
)

plot(allEffects(mod_FDivergence_fac), type="response")
summary(mod_FDivergence_num)
car::Anova(mod_FDivergence_num)


# numeric year model
df_FDivergence_num <- as.data.frame(allEffects(mod_FDivergence_num)[[5]])
df_FDivergence_num$model <- "numeric"

# factor year model
df_FDivergence_fac <- as.data.frame(allEffects(mod_FDivergence_fac)[[5]])
df_FDivergence_fac$model <- "factor"
df_FDivergence_fac$year_num <- as.numeric(as.character(df_FDivergence_fac$year))



ggplot() +
  # --- NUMERIC MODEL (smooth trend) ---
  geom_line(data = df_FDivergence_num,
            aes(x = year_num, y = fit, colour = restauration),
            size = 1) +

  geom_ribbon(data = df_FDivergence_num,
              aes(x = year_num, ymin = lower, ymax = upper, fill = restauration),
              alpha = 0.2) +

  # --- FACTOR MODEL (year-specific points) ---
  geom_point(data = df_FDivergence_fac,
             aes(x = year_num, y = fit, colour = restauration),
             size = 2) +
  geom_area() + scale_x_continuous(breaks=min(glmm_df$year):max(glmm_df$year))+
  geom_errorbar(data = df_FDivergence_fac,
              aes(x = year_num, ymin = lower, ymax = upper, colour = restauration),
              width = 0.2)+
  
  labs(x = "Year",
       y = "Predicted FDivergence",
       colour = "Restoration",
       fill = "Restoration") +
  theme_classic()

simulationOutput <- simulateResiduals(mod_FDivergence_num, plot = TRUE)
```
```{r}
#RaoQ GLMM####

#numeric
mod_RaoQ_num <- glmmTMB(
  RaoQ ~ log(surface_water + 1) + year_num * restauration +
    cloudiness + wind + julian_day, #(1|site/Subsite),
  family = gaussian(),
  data = fd_df
)

#factor
mod_RaoQ_fac <- glmmTMB(
  RaoQ ~ log(surface_water + 1) + year * restauration +
    cloudiness + wind + julian_day, #(1|site/Subsite),
  family = gaussian(),
  data = fd_df
)

plot(allEffects(mod_RaoQ_fac), type="response")
summary(mod_RaoQ_num)
car::Anova(mod_RaoQ_num)


# numeric year model
df_RaoQ_num <- as.data.frame(allEffects(mod_RaoQ_num)[[5]])
df_RaoQ_num$model <- "numeric"

# factor year model
df_RaoQ_fac <- as.data.frame(allEffects(mod_RaoQ_fac)[[5]])
df_RaoQ_fac$model <- "factor"
df_RaoQ_fac$year_num <- as.numeric(as.character(df_RaoQ_fac$year))



ggplot() +
  # --- NUMERIC MODEL (smooth trend) ---
  geom_line(data = df_RaoQ_num,
            aes(x = year_num, y = fit, colour = restauration),
            size = 1) +

  geom_ribbon(data = df_RaoQ_num,
              aes(x = year_num, ymin = lower, ymax = upper, fill = restauration),
              alpha = 0.2) +

  # --- FACTOR MODEL (year-specific points) ---
  geom_point(data = df_RaoQ_fac,
             aes(x = year_num, y = fit, colour = restauration),
             size = 2) +
  geom_area() + scale_x_continuous(breaks=min(glmm_df$year):max(glmm_df$year))+
  geom_errorbar(data = df_RaoQ_fac,
              aes(x = year_num, ymin = lower, ymax = upper, colour = restauration),
              width = 0.2)+
  
  labs(x = "Year",
       y = "Predicted RaoQ",
       colour = "Restoration",
       fill = "Restoration") +
  theme_classic()

simulationOutput <- simulateResiduals(mod_RaoQ_num, plot = TRUE)
```

```{r}
plot(PCOA1, PCOA2, pch = 19)
text(PCOA1, PCOA2,
     labels = rownames(traits_matrix),
     cex = 0.6, pos = 3)
```
```{r}
#PLOT PREDICTED ABUNDANCE AGAINST EVENNESS####
# Evenness predictions (on response scale!)
glmm_df$pred_evenness <- predict(mod_eveness_num, newdata = glmm_df, type = "response")

# Abundance predictions
glmm_df$pred_abundance <- predict(mod_abundance_num, newdata = glmm_df, type = "response")
```

